---
title: "02_enrichment_A3_FS"
author: "Fabrice Schlatter"
date: "17.11.2021"
output: html_document
---

#aim: do the enrichment based on 01_direct_processing_X outputs
#based on the RVL Rscript and also does the downstream excel analysis which were used for the original DMS enrichement analysis
#runtime: fast


#load libraries
```{r, warning=FALSE, message=FALSE}
library(Biostrings)
library(tidyverse)
```


#choose sample and sample overview
to set path and define sample names
```{r}
##choose sample, set paths and give it a run description

path <- "C:/Users/" #make sure the path ends with a /
path <- '/Users/'

setwd(path)
#list.files(path)

#Genewiz folder
files <- sort(list.files(paste0(path, "01_direct_processing_of_fasta/output/"), pattern="counts", full.names = TRUE))
files

#extract sample names, assuming / as delimiter
sample.names <- sapply(strsplit(basename(files), "/"), `[`, 1)
sample.names
sample.names %>% length()

#set indices for which samples, match them to read in
#doublecheck always that it matches!
i1 <- 1
i2 <- 2
i3 <- 3

print(paste(sample.names[i1], sample.names[i2], sample.names[i3]))

#read in
plasmid <- read.csv(files[i1])
coculture <- read.csv(files[i2])
binding <- read.csv(files[i3])

#sample name / title
sample_name <- sapply(strsplit(sample.names[i1], "-"), `[`, 1)
sample_name
#sample name manual renaming
sample_name <- "a305"
sample_name

```


```{r}
#global variables
#add wt sequence of the mutagenesis region here

aa_wt <- AAString("ASSPNMADEQY")
aa_wt <- AAString("IQSSPGGAGSYQLT")

aa_wt

```


#boil down to freqencies per aa position

###checkpoint###
the next three chunks have to be run one by one for each sample (here: plasmid, coculture and binding). you do this by running only one out of the three lines in this next chunk and thus select which data should be saved in temp. then you run the entire second chunk based on that temp variable. then in the third chunks you again only run those lines that correspond to the right dataset and save it this way to a new variable. then you can go back to here, overwrite temp anew and continue as before.
###

#chunk1
```{r}
stop()

#choose input, one out of three
temp <- plasmid
temp <- coculture
temp <- binding

stop()
```

#chunk2
```{r}
#reshape & quick overview, wt fraction
temp_aa <- data.frame(aa = AAStringSet(temp$aa), n = temp$n)
print("wt count:")
wt_count <- temp_aa %>% filter(.,as.character(aa)==as.character(aa_wt)) %>% .$n %>% sum()
wt_count
print("wt freq:")
wt_freq <- wt_count / sum(temp_aa$n)
wt_freq
#unique function was called on DNA seq and not AA seq: obviously there is still some on AA:
table(temp_aa$aa) %>% sort(., decreasing = TRUE) %>% head(.,6)
#merge them:
temp_aa <- aggregate(n ~ aa,temp_aa, sum)
temp_aa <- temp_aa[order(temp_aa$n, decreasing = TRUE),]

#option to truncate (discarding info vs robustness against seq errors)
temp_aa_t <- subset(temp_aa, n > 5)

#clean up
#a lot happened already in the processing pipeline e.g. based on length: (not necessary here anymore)
nchar(temp_aa_t$aa) %>% table()
#remove fuzzy X
ind <- vcountPattern("X", temp_aa_t$aa, max.mismatch = 0)
sum(ind)
temp_aa_t <- temp_aa_t[which(as.logical(!ind)),]
#could exclude * here too, but then adjust single_muts_missed in the loop below
#actually dont do it because this is real info if it passes all other filters in the pipeline so far (ie real mRNA)

#add freqency
temp_aa_t$freq <- temp_aa_t$n / sum(temp_aa_t$n)
sum(temp_aa_t$freq)

#single mutations only (DMS) - (different to RVL)
ind <- vcountPattern(aa_wt, temp_aa_t$aa, max.mismatch = 1)
temp_aa_ts <- temp_aa_t[which(as.logical(ind)),]
print("multi muts dropped:")
sum(ind==0)
print("single mut ratio:")
sum(temp_aa_ts$freq) #ie while I lost a lot of unique seqs, they all had very low freq
#readjust frequency - (different to RVL)
temp_aa_ts$freq <- temp_aa_ts$n / sum(temp_aa_ts$n)
sum(temp_aa_ts$freq)
print("wt count readj:")
wt_count <- temp_aa_ts %>% filter(.,as.character(aa)==as.character(aa_wt)) %>% .$n %>% sum()
wt_count
print("wt freq readj:")
wt_freq <- wt_count / sum(temp_aa_ts$n)
wt_freq

#split aa
temp_aa_tss <- strsplit(temp_aa_ts$aa, split = "") %>% as.data.frame %>% t %>% cbind(temp_aa_ts,.)


#get the point mutations per position
#this is based on the fact that we only have data with single point mutations now

#start with wt
temp_wt <- data.frame(aa = as.character(aa_wt),
                      n = wt_count,
                      freq = wt_freq,
                      a = strsplit(as.character(aa_wt), split = ""),
                      pos = c(1:nchar(aa_wt)))
names(temp_wt)[4] <- "a"

#append all other positions, loop over each pos, count if we see all single muts
a_ref = strsplit(as.character(aa_wt), split = "")
single_muts_missed <- length(a_ref[[1]])*20 #20 = 19aa (instead of the wt aa) + * (stop)
for (pos in 1:nchar(aa_wt)){
    #get info
    temp_pos <- subset(temp_aa_tss, temp_aa_tss[pos+3] != a_ref[[1]][pos])
    #print(paste0((nrow(temp_pos) == length(unique(temp_pos[,pos+3]))), nrow(temp_pos), length(unique(temp_pos[,pos+3]))))#check against double muts, must be equal/true
    single_muts_missed <- single_muts_missed - nrow(temp_pos)
    temp_pos <- temp_pos[,c(1,2,3,(pos+3))]
    names(temp_pos)[4] <- "a"
    temp_pos$pos <- pos
    #append
    temp_wt <- rbind(temp_wt, temp_pos)
}
print("single_muts_missed:")
single_muts_missed

#check on percentages
#sum(temp_wt$freq)-sum(temp_wt$freq[1:14])+temp_wt$freq[1]
#this is how they sum to 1, it is basically distorted by 13*wt_freq, as it was only accounted for once (13+1)

##order nicely and according to heatmap design below, span matrix based on freq, a and pos
#(however, ggplot seems to need the long df as input)
temp_complete <- temp_wt[,c(4,5,3)]
#change/add level order
x_level_order <- as.character(c(1:nchar(aa_wt))) #length of CDR3b
y_level_order <- c("A", "C", "D", "E", "F", "G", "H", "I", "K", "L", "M", "N", "P", "Q", "R", "S", "T", "V", "W", "Y", "*")
#apply
temp_complete$a <- factor(temp_complete$a, levels = y_level_order)
temp_complete$pos <- factor(temp_complete$pos, levels = x_level_order)
temp_complete <- temp_complete[order(temp_complete$a, temp_complete$pos),]
#wider format ie matrix
temp_matrix <- pivot_wider(temp_complete, names_from = pos, values_from = freq)
temp_matrix[is.na(temp_matrix)] <- 0 #set NA to 0
#if A is not complete, pivotwider appends missing at the end...ie need to order the matrix again (also to be safe)
temp_matrix <- temp_matrix[, c("a",x_level_order)]
row.names(temp_matrix) <- temp_matrix$a
temp_matrix <- temp_matrix[y_level_order,]
#transform back into long format so that we have all the 0 entries included (RVL set NA to 0)
temp_complete_0 <- pivot_longer(temp_matrix, cols = c(2:(nchar(aa_wt)+1)))
names(temp_complete_0) <- names(temp_complete)
#should be sorted, but to be safe
temp_complete_0$a <- factor(temp_complete_0$a, levels = y_level_order)
temp_complete_0$pos <- factor(temp_complete_0$pos, levels = x_level_order)
temp_complete_0 <- temp_complete_0[order(temp_complete_0$a, temp_complete_0$pos),]


```

#chunk3
```{r}
stop()

#save output of chuck to new vairables, one out of three
df_plasmid <- temp_complete_0
m_plasmid <- temp_matrix

df_coculture <- temp_complete_0
m_coculture <- temp_matrix

df_binding <- temp_complete_0
m_binding <- temp_matrix

stop()
```

###checkpoint###
at this point you should have run all three samples (here: plasmid, coculture and binding) seperately through the above three chunks.only then you can contine without any errors :)
###

#csv output
for degeneratecodon package and or excel/prism heatmaps
```{r}
#create output
path_out <- paste0(getwd(), "/output/")
setwd(path_out)

#addon for eg threshold
add_on <- "_t5"

#output
write.csv(df_plasmid, file = paste0(sample_name, add_on, "_df_plasmid", ".csv"), row.names = FALSE)
write.csv(m_plasmid, file = paste0(sample_name, add_on, "_m_plasmid", ".csv"), row.names = FALSE)

write.csv(df_coculture, file = paste0(sample_name, add_on, "_df_coculture", ".csv"), row.names = FALSE)
write.csv(m_coculture, file = paste0(sample_name, add_on, "_m_coculture", ".csv"), row.names = FALSE)

write.csv(df_binding, file = paste0(sample_name, add_on, "_df_binding", ".csv"), row.names = FALSE)
write.csv(m_binding, file = paste0(sample_name, add_on, "_m_binding", ".csv"), row.names = FALSE)

```


#-------------

#dms analysis
compare the freq matrixes and create a heatmap output (prep for enrichment analysis)
```{r}
#plot heatmaps
#rectangles: https://stackoverflow.com/questions/13258454/marking-specific-tiles-in-geom-tile-geom-raster

#specify which heatmap to plot, based on long df format
toplot <- df_coculture

##auxiliaries
#level order for axis
x_level_order <- as.character(c(1:14))
y_level_order <- rev(c("A", "C", "D", "E", "F", "G", "H", "I", "K", "L", "M", "N", "P", "Q", "R", "S", "T", "V", "W", "Y", "*"))
#order
toplot$a <- factor(toplot$a, levels = y_level_order)
toplot$pos <- factor(toplot$pos, levels = x_level_order)
#which get a frame (=wt)
toplot$frames <- FALSE
for (pos in 1:nchar(aa_wt)){
  temp <- a_ref[[1]][pos]
  toplot$frames[which(toplot$a == temp & toplot$pos == pos)] <- TRUE
}
#aux df
frames <- toplot[toplot$frames, c("a", "pos")]
frames$a <- as.integer(frames$a)
frames$pos <- as.integer(frames$pos)


#plot
p1 = ggplot(data = toplot) +
     geom_tile(aes(x = pos,
                   y = a,
                   fill = log(freq)), color = "white") +
     scale_fill_gradient2(low = "blue", high = "red", mid = "white", midpoint = -5, na.value = "grey50", name = "log(Freq)") +
     geom_rect(data=frames, size=1, fill=NA, colour="black", 
               aes(xmin = pos-0.5, xmax = pos+0.5, ymin = a-0.5, ymax = a+0.5)) +
    ggtitle("DMS Distribution") + xlab("Position") + ylab("Amino Acid") +
    coord_equal() + theme_minimal()
p1

#non-log
p1 = ggplot(data = toplot) +
     geom_tile(aes(x = pos,
                   y = a,
                   fill = freq), color = "white") +
     scale_fill_gradient2(low = "blue", high = "red", mid = "white", midpoint = 0.1, na.value = "grey50", name = "Freq") +
     geom_rect(data=frames, size=1, fill=NA, colour="black", 
               aes(xmin = pos-0.5, xmax = pos+0.5, ymin = a-0.5, ymax = a+0.5)) +
    ggtitle("DMS Distribution") + xlab("Position") + ylab("Amino Acid") +
    coord_equal() + theme_minimal()
p1


```

```{r}
#create output
path_out <- paste0(getwd(), "/output/")
setwd(path_out)

tiff(filename = paste0(sample_name, add_on, "_plasmid", "_freq.tiff"), units="in", width=5, height=5, res=300)
p1
dev.off()

```


#-------------

#enrichment analysis
compare the different matrixes and create a heatmap output
```{r}
#plot heatmaps
#rectangles: https://stackoverflow.com/questions/13258454/marking-specific-tiles-in-geom-tile-geom-raster

ncol <- 12

#specify which enrichment to plot, based on wide matrix format
sample_toplot <- m_binding #based on a selection typically
normalize_toplot <- m_plasmid #df_plasmid typically

#normalize
normed_toplot <- sample_toplot
normed_toplot[,2:ncol] <- sample_toplot[,2:ncol]/normalize_toplot[,2:ncol]

#covert back to long df format for ggplot
toplot <- pivot_longer(normed_toplot, cols = c(2:ncol))
names(toplot) <- c("a", "pos", "freq")

##auxiliaries
#level order for axis
x_level_order <- as.character(c(1:14))
y_level_order <- rev(c("A", "C", "D", "E", "F", "G", "H", "I", "K", "L", "M", "N", "P", "Q", "R", "S", "T", "V", "W", "Y", "*"))
#order
toplot$a <- factor(toplot$a, levels = y_level_order)
toplot$pos <- factor(toplot$pos, levels = x_level_order)
#which get a frame (=wt)
toplot$frames <- FALSE
for (pos in 1:nchar(aa_wt)){
  temp <- a_ref[[1]][pos]
  toplot$frames[which(toplot$a == temp & toplot$pos == pos)] <- TRUE
}
#aux df
frames <- toplot[toplot$frames, c("a", "pos")]
frames$a <- as.integer(frames$a)
frames$pos <- as.integer(frames$pos)


#plot
p2 = ggplot(data = toplot) +
     geom_tile(aes(x = pos,
                   y = a,
                   fill = log(freq)), color = "white") +
     scale_fill_gradient2(low = "blue", high = "red", mid = "white", midpoint = -1, na.value = "grey50", name = "log(ENR)") +
     geom_rect(data=frames, size=1, fill=NA, colour="black", 
               aes(xmin = pos-0.5, xmax = pos+0.5, ymin = a-0.5, ymax = a+0.5)) +
    ggtitle("Enrichment") + xlab("Position") + ylab("Amino Acid") +
    coord_equal() + theme_minimal()
p2

#non-log
p2 = ggplot(data = toplot) +
     geom_tile(aes(x = pos,
                   y = a,
                   fill = freq), color = "white") +
     scale_fill_gradient2(low = "blue", high = "red", mid = "white", midpoint = 8, na.value = "grey50", name = "ENR") +
     geom_rect(data=frames, size=1, fill=NA, colour="black", 
               aes(xmin = pos-0.5, xmax = pos+0.5, ymin = a-0.5, ymax = a+0.5)) +
    ggtitle("Enrichment") + xlab("Position") + ylab("Amino Acid") +
    coord_equal() + theme_minimal()
p2


```

```{r}
#create output
path_out <- paste0(getwd(), "/output/")
setwd(path_out)

#addon for eg legend
add_on2 <- ""

tiff(filename = paste0(sample_name, add_on, add_on2, "_binding", "_enrichment.tiff"), units="in", width=5, height=5, res=300)
p2
dev.off()

```


```{r}
stop()
```


```{r}
#sandbox



```


